---
title: "Health outcomes in Bulgaria: simulated effects of obesogenic environmental changes in adulthood versus childhood"
subtitle: "Extrapolate BMI classes beyond observed time and create BMI transition matrices"
output: htlm_document
author: Elena Milkovska, Pieter van Baal
BMI extrapolation methodology: https://data.princeton.edu/eco572/leecarter
transition matrix methodology: van de Kassteele J, Hoogenveen RT, Engelfriet PM, van Baal PHM, Boshuizen HC (2012). Estimating net transition probabilities from cross-sectional data with application to risk factors in chronic disease modeling. Statistics in Medicine 31(6), 533-543. http://onlinelibrary.wiley.com/doi/10.1002/sim.4423/abstract 
transition matrix code: https://github.com/kassteele/Net-transition-probabilities
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
library("pacman")
p_load("mgcv","readxl","dplyr","ggplot2","gamlss","tseries","forecast","tidyverse","Matrix","lpSolve")


# Load file, define working directory and prediction period  
wd            <- "C:/Example directory/"                     # Input the main directory containing the files.
bmi           <- read_excel(paste(wd,"data/source/NCD Risc BMI/NCD_RisC_Lancet_2017_BMI_prevalence.xlsx",sep=""))
country       <- "Bulgaria"
v_years_ahead <- 2016:2030       # For which years to predict BMI classes
last_year     <- 2016            # Last year available of actual data   
age.pred      <- 20:100          # For which ages
input_year    <- 2022            # For which year to generate BMI transition matrices 

# Define prediction data frame
pred          <- data.frame(country = country,
                            year = rep(v_years_ahead,each=length(unique(bmi$age_group))),
                            age_group = unique(bmi$age_group),
                            matrix(NA,ncol = 6))
names(pred)   <- c("country","year","age_group","ow.fit","ow.low","ow.high","ob.fit","ob.low","ob.high") 
pred          <- rbind(pred,pred)
pred$sex      <- rep(c("Female","Male"),each=dim(pred)[1] * 0.5)
```

#01. Extrapolating BMI prevalence beyond 2016

The Lee–Carter model is a numerical algorithm used in mortality forecasting and life expectancy forecasting. But the method can also be used to other values, like prevalence of BMI. The input to the model is a matrix of age specific mortality rates ordered monotonically by time, usually with ages in columns and years in rows.

##01.01. Analysis for Women - Overweight category

```{r 01.01. Input}
df <- bmi[bmi$country == country & bmi$sex== "Female",c("year","age_group","ow")]
df <- pivot_wider(df,names_from=age_group,values_from = ow)
head(df)
```

### SDV decomposition - Women

```{r 01.01. Rework default data}
# Converт to OR and the ln(OR)

ncol        <- dim(df)[2]
df[,2:ncol] <- lapply(df[,2:ncol], function(x) x/(1-x))
df[,2:ncol] <- log(df[,2:ncol])

# Calculate 'a_x' vector. The average log(OR) over time for each age group. 
a <- colMeans(df[,2:ncol])

# De-mean the matrix - log(m_xt)
for (j in 2:ncol) df[,j] <- df[,j] - a[j-1]

# Calculate 'b_x' and 'k_t' vectors 
M <- as.matrix(df[,2:ncol])
d <- svd(M, 1, 1)
b <- d$v/sum(d$v)
k <- d$u * sum(d$v) * d$d[1]

trend <- data.frame(year = c(unique(bmi$year),v_years_ahead[-1]), k = c(k,rep(NA,length(v_years_ahead)-1)))
```

### ARIMA Modelling of k - Women

```{r 01.01. ARIMA for k}
#Test for stationarity 
tsK <- ts(k, start = 1975, frequency = 1)
acf(tsK) 
adf.test(tsK)                        # Does no appear to stationary - ADF test has a high p-value, indicating that the time-series is not stationary. 
                                     # Same can be observed from the gradually decaying sample ACF.
#Make stationary  
tssK <- diff(tsK, differences=1)
adf.test(tssK)                       # Stationary after the first.

fit0 <- auto.arima(tsK,trace = TRUE) # Picks ARIMA(1,2,0)
checkresiduals(fit0)                 # Large p-value. Thus, we can conclude that the residuals are not distinguishable from a white noise series. 
                                     # Residuals' mean is around 0, even though there's 1 significant spike the ACf plot.
```

```{r 01.01. Predict k}
z   <- qnorm(0.975)
se  <- sqrt(fit0$sigma2)

trend$k[trend$year > last_year]      <- as.vector(predict(fit0,n.ahead = length(v_years_ahead)-1)$pred)
trend$k.low[trend$year > last_year]  <-  trend$k[trend$year > last_year] - z*se
trend$k.high[trend$year > last_year] <-  trend$k[trend$year > last_year] + z*se

#Check how well estimates fit for the last year of actual data - 2016. 
x <- exp(a + b * trend$k[trend$year==last_year])
x <- x/(1+x)
bmi[bmi$country== country & bmi$year == last_year & bmi$sex == "Female","ow.fit"] <- x


error <- bmi[bmi$country == country & bmi$sex == "Female" & bmi$year == last_year,c("ow","ow.fit")]
error <- (error$ow - error$ow.fit)^2
sum(error) #Sum of squared errors is close to 0
```

```{r 01.01. Predict ow values}

for (t in 1:length(v_years_ahead)) {
  x <- exp(a + b * trend$k[trend$year==v_years_ahead[t]])
  x <- x/(1+x)
  pred$ow.fit[pred$year == v_years_ahead[t] & pred$sex == "Female"] <-  x
  x <- exp(a + b * trend$k.low[trend$year==v_years_ahead[t]])
  x <- x/(1+x)
  pred$ow.low[pred$year == v_years_ahead[t] & pred$sex == "Female"] <- x
  x <- exp(a + b * trend$k.high[trend$year==v_years_ahead[t]])
  x <- x/(1+x)
  pred$ow.high[pred$year == v_years_ahead[t] & pred$sex == "Female"] <- x
}

trend_plot <- trend
names(trend_plot) <- c("year","ow_Female")
trend_plot[,3:4] <- NULL
```


## 01.02. Analysis for Women - Obesity category


```{r 01.02. Women ob.fit}
df <- bmi[bmi$country == country & bmi$sex== "Female",c("year","age_group","ob")]
df <- pivot_wider(df,names_from=age_group,values_from = 'ob')

#Convert to OR and the ln(OR)

ncol <- dim(df)[2]
df[,2:ncol]  <- lapply(df[,2:ncol], function(x) x/(1-x))
df[,2:ncol]  <- log(df[,2:ncol])

#Calculate 'a_x' vector. The average log(mortality) over time for each age group. 
a <- colMeans(df[,2:ncol])

#De-mean the matrix - log(m_xt)
for (j in 2:ncol) df[,j] <- df[,j] - a[j-1]

#Calculate 'b_x' and 'k_t' vectors 
M <- as.matrix(df[,2:ncol])
d <- svd(M, 1, 1)
b <- d$v/sum(d$v)
k <- d$u * sum(d$v) * d$d[1]

#Plot k
trend <- data.frame(year = c(unique(bmi$year),v_years_ahead[-1]), k = c(k,rep(NA,length(v_years_ahead)-1)))
ggplot(trend, aes(year, k)) + geom_line() 

tsK <- ts(k, start = 1975, frequency = 1)
adf.test(diff(tsK, differences=1))

fit0 <- auto.arima(tsK,trace = TRUE) # picks ARIMA(1,2,0)
checkresiduals(fit0)                 # Large p-value. Thus, we can conclude that the residuals are not distinguishable from a white noise series.   
                                     # Residuals' mean is around 0, 1rst spike is significant on the ACF plot.
z <- qnorm(0.975)
se <- sqrt(fit0$sigma2)
trend$k[trend$year > last_year] <- as.vector(predict(fit0,n.ahead = length(v_years_ahead)-1)$pred)
trend$k.low[trend$year > last_year] <-  trend$k[trend$year > last_year] - z*se
trend$k.high[trend$year > last_year] <-  trend$k[trend$year > last_year] + z*se

#Check how well estimates fit for the last year of actual data - 2016. 
x <- exp(a + b * trend$k[trend$year==last_year])
x <- x/(1+x)
bmi[bmi$country== country & bmi$year == last_year & bmi$sex == "Female","ob.fit"] <- x

ggplot(bmi[bmi$country== country & 
          bmi$year == last_year & bmi$sex == "Female",],aes(age_group)) +
  geom_point(aes(y=ob),col="blue") +
  geom_point(aes(y=ob.fit))


error <- bmi[bmi$country == country & bmi$sex == "Female" & bmi$year == last_year,c("ob","ob.fit")]
error <- (error$ob - error$ob.fit)^2
sum(error) #Sum of squared errors is 0 

for (t in 1:length(v_years_ahead)) {
  x <- exp(a + b * trend$k[trend$year==v_years_ahead[t]])
  x <- x/(1+x)
  pred$ob.fit[pred$year == v_years_ahead[t] & pred$sex == "Female"] <-  x
  x <- exp(a + b * trend$k.low[trend$year==v_years_ahead[t]])
  x <- x/(1+x)
  pred$ob.low[pred$year == v_years_ahead[t] & pred$sex == "Female"] <- x
  x <- exp(a + b * trend$k.high[trend$year==v_years_ahead[t]])
  x <- x/(1+x)
  pred$ob.high[pred$year == v_years_ahead[t] & pred$sex == "Female"] <- x
}

trend_plot$ob_Female <- trend$k
```

##01.03. Analysis for Men - Overweight category

```{r 01.03. Men ow.fit}
df <- bmi[bmi$country == country & bmi$sex== "Male",c("year","age_group","ow")]
df <- pivot_wider(df,names_from=age_group,values_from = 'ow')

#Convert to OR and the ln(OR)

ncol <- dim(df)[2]
df[,2:ncol]  <- lapply(df[,2:ncol], function(x) x/(1-x))
df[,2:ncol] <- log(df[,2:ncol])

#Calculate 'a_x' vector. The average log(mortality) over time for each age group. 
a <- colMeans(df[,2:ncol])

#De-mean the matrix - log(m_xt)
for (j in 2:ncol) df[,j] <- df[,j] - a[j-1]

#Calculate 'b_x' and 'k_t' vectors 
M <- as.matrix(df[,2:ncol])
d <- svd(M, 1, 1)
b <- d$v/sum(d$v)
k <- d$u * sum(d$v) * d$d[1]

#Plot k
trend <- data.frame(year = c(unique(bmi$year),v_years_ahead[-1]), k = c(k,rep(NA,length(v_years_ahead)-1)))
ggplot(trend, aes(year, k)) + geom_line() 

tsK <- ts(k, start = 1975, frequency = 1)
adf.test(diff(tsK, differences=1))

fit0 <- auto.arima(tsK,trace = TRUE) # picks ARIMA(1,2,0)
checkresiduals(fit0)                 # Large p-value. Thus, we can conclude that the residuals are not distinguishable from a white noise series. 
                                     # Residuals' mean is around 0, 1rst spike is significant on the ACF plot.
z <- qnorm(0.975)
se <- sqrt(fit0$sigma2)
trend$k[trend$year > last_year]      <- as.vector(predict(fit0,n.ahead = length(v_years_ahead)-1)$pred)
trend$k.low[trend$year > last_year]  <-  trend$k[trend$year > last_year] - z*se
trend$k.high[trend$year > last_year] <-  trend$k[trend$year > last_year] + z*se

#Check how well estimates fit for the last year of actual data - 2016. 
x <- exp(a + b * trend$k[trend$year==2016])
x <- x/(1+x)
bmi[bmi$country==country & bmi$year == last_year & bmi$sex == "Male","ow.fit"] <- x

ggplot(bmi[bmi$country==country & 
          bmi$year == last_year & bmi$sex == "Male",],aes(age_group)) +
  geom_point(aes(y=ow),col="blue") +
  geom_point(aes(y=ow.fit))


error <- bmi[bmi$country == country & bmi$sex == "Male" & bmi$year == last_year,c("ow","ow.fit")]
error <- (error$ow - error$ow.fit)^2
sum(error) #Sum of squared errors is 0 

for (t in 1:length(v_years_ahead)) {
  x <- exp(a + b * trend$k[trend$year==v_years_ahead[t]])
  x <- x/(1+x)
  pred$ow.fit[pred$year == v_years_ahead[t] & pred$sex == "Male"] <-  x
  x <- exp(a + b * trend$k.low[trend$year==v_years_ahead[t]])
  x <- x/(1+x)
  pred$ow.low[pred$year == v_years_ahead[t] & pred$sex == "Male"] <- x
  x <- exp(a + b * trend$k.high[trend$year==v_years_ahead[t]])
  x <- x/(1+x)
  pred$ow.high[pred$year == v_years_ahead[t] & pred$sex == "Male"] <- x
}

trend_plot$ow_Male <- trend$k
```


##01.04. Analysis for Men - Obesity category

```{r Men ob.fit}
df <- bmi[bmi$country == country & bmi$sex== "Male",c("year","age_group","ob")]
df <- pivot_wider(df,names_from=age_group,values_from = 'ob')

#Convert to OR and the ln(OR)

ncol <- dim(df)[2]
df[,2:ncol]  <- lapply(df[,2:ncol], function(x) x/(1-x))
df[,2:ncol] <- log(df[,2:ncol])

#Calculate 'a_x' vector. The average log(mortality) over time for each age group. 
a <- colMeans(df[,2:ncol])

#De-mean the matrix - log(m_xt)
for (j in 2:ncol) df[,j] <- df[,j] - a[j-1]

#Calculate 'b_x' and 'k_t' vectors 
M <- as.matrix(df[,2:ncol])
d <- svd(M, 1, 1)
b <- d$v/sum(d$v)
k <- d$u * sum(d$v) * d$d[1]

#Plot k
trend <- data.frame(year = c(unique(bmi$year),v_years_ahead[-1]), k = c(k,rep(NA,length(v_years_ahead)-1)))
ggplot(trend, aes(year, k)) + geom_line()

tsK <- ts(k, start = 1975, frequency = 1)
adf.test(diff(tsK, differences=1))

fit0 <- auto.arima(tsK,trace = TRUE) #picks ARIMA(1,2,0)
checkresiduals(fit0) #Large p-value. Thus, we can conclude that the residuals are not distinguishable from a white noise series. Residuals' mean is around 0, 1rst spike is significant on the ACF plot.

z <- qnorm(0.975)
se <- sqrt(fit0$sigma2)
trend$k[trend$year > last_year]      <- as.vector(predict(fit0,n.ahead = length(v_years_ahead)-1)$pred)
trend$k.low[trend$year > last_year]  <-  trend$k[trend$year > last_year] - z*se
trend$k.high[trend$year > last_year] <-  trend$k[trend$year > last_year] + z*se

#Check how well estimates fit for the last year of actual data - 2016. 
x <- exp(a + b * trend$k[trend$year==last_year])
x <- x/(1+x)
bmi[bmi$country==country & bmi$year == last_year & bmi$sex == "Male","ob.fit"] <- x

ggplot(bmi[bmi$country== country & 
          bmi$year == last_year & bmi$sex == "Male",],aes(age_group)) +
  geom_point(aes(y=ob),col="blue") +
  geom_point(aes(y=ob.fit))


error <- bmi[bmi$country == country & bmi$sex == "Male" & bmi$year == last_year,c("ob","ob.fit")]
error <- (error$ob - error$ob.fit)^2
sum(error) #Sum of squared errors is 0 

for (t in 1:length(v_years_ahead)) {
  x <- exp(a + b * trend$k[trend$year==v_years_ahead[t]])
  x <- x/(1+x)
  pred$ob.fit[pred$year == v_years_ahead[t] & pred$sex == "Male"] <-  x
  x <- exp(a + b * trend$k.low[trend$year==v_years_ahead[t]])
  x <- x/(1+x)
  pred$ob.low[pred$year == v_years_ahead[t] & pred$sex == "Male"] <- x
  x <- exp(a + b * trend$k.high[trend$year==v_years_ahead[t]])
  x <- x/(1+x)
  pred$ob.high[pred$year == v_years_ahead[t] & pred$sex == "Male"] <- x
}

trend_plot$ob_Male <- trend$k
```


#02 Smooth out BMI prevalences across age. 1-year age groups needed


```{r Get the GAM with smallest GCV}
pred_age_groups <- bmi <- pred
bmi$age_group[bmi$age_group == "85plus"] <- "85-100"
v_sex            <- c("Female","Male")
l_names          <- strsplit(bmi$age_group,split = "-")
l_names          <- array(unlist(l_names), dim=c(1, 2, dim(bmi)[1]))
bmi[,c("start")] <- l_names[1,1,]
bmi[,c("end")]   <- l_names[1,2,]
bmi$start        <- as.numeric(bmi$start)
bmi$end          <- as.numeric(bmi$end)

for (i in 1:dim(bmi)[1]) {
  range      <- c(bmi$start[i],bmi$end[i])
  bmi$age[i] <- median(range)
}

bmi$age      <- as.numeric(bmi$age)
obmi         <- bmi
obmi[,c("ow.fit","ob.fit")] <- lapply(obmi[,c("ow.fit","ob.fit")],function(x) x/(1-x))

#Smooth out overweight - "gp" for OW Women & Men
counter = 0
for (b in c("ow.fit","ob.fit")) {
 for (s in 1:2) {
  for (t in 1:length(v_years_ahead)) {
    counter         <- counter + 1
    df              <- obmi[obmi$year==v_years_ahead[t] & obmi$sex==v_sex[s],]
    fit             <- list()
    bs_type         <- c("tp","ts","ds","cr","cs","ps","re","gp")
    fit_gam         <- vector(mode = "list", length = length(bs_type))
    names(fit_gam)  <- bs_type
    gcv             <- rep(0,length(bs_type)) 
  
    for (i in 1:length(bs_type)) {
      formula_gam   <- as.formula(paste0('log(',b,') ~ s(age,bs="',bs_type[i],'")',sep=''))
      fit_gam[[i]]  <- gam(formula_gam, data=df)
      gcv[i]        <- fit_gam[[i]]$gcv.ubre
      }

    pred.df         <- data.frame(year = v_years_ahead[t],sex = v_sex[s],age = age.pred)
    min_gcv         <- fit_gam[[bs_type[which(gcv == min(gcv))[1]]]]
    pred.df$fit     <- exp(predict(min_gcv, newdata = pred.df))
    pred.df$log.se  <- predict(min_gcv, newdata = pred.df,se.fit = TRUE)$se.fit
    pred.df$fit     <- pred.df$fit/(1+pred.df$fit)
    pred.df$bmi     <- b
    
    if (counter==1) {
      predictions   <- pred.df
      } else {
      predictions   <- rbind(predictions,pred.df)
     }  
  }
 }
}

pred        <- predictions
pred$bmi    <- sub(".fit","",pred$bmi)
pred        <- pred %>% pivot_wider(names_from = bmi, values_from = c(fit,log.se))
names(pred) <- c("year","sex","age","ow","ob","ow.OR.se","ob.OR.se")
pred$no     <- 1 - (pred$ow + pred$ob)
```

```{r Save files}
write.csv(pred,file=paste(wd,"data/NCD_RisC_Lancet_smooth_2030_Bulgaria.csv",sep = ""))
write.csv(trend_plot,file=paste(wd,"data/K-value_BMI.csv",sep = ""))
write.csv(pred_age_groups,file=paste(wd,"data/pred_age_groups.csv",sep = ""))
```


#03 Creating BMI transition matrices - Linear Programming

Run this same code both for men and women separately. 

##03.01 Women, 1-year age BMI prevalences taken from year 2022.

```{r Import Data - Use 2022 only}
df        <- read_csv(paste(wd,"data/NCD_RisC_Lancet_smooth_2030_Bulgaria.csv",sep = ""))
df        <- df[df$year == input_year & df$sex == "Female",c("age","no","ow","ob")]
df[,2:4]  <- round(df[,2:4] * 1000)
names(df) <- c("age","normal","overweight","obese")
bmi.data  <- df
head(bmi.data)
```

```{r Fit multinomial P-splines}
# Number of classes
K <- 3

# Reshape bmi.data into person specific records (long format)
# We need this for the multinomial P-spline fit procedure
# 0 = normal, 1 = overweight, 2 = obese
bmi.data.long <- with(bmi.data, data.frame(
  age = rep(rep(age, times = K), times = c(normal, overweight, obese)),
  y = rep(0:(K - 1), times = c(sum(normal), sum(overweight), sum(obese)))))

# Fit multinomial P-splines
# See help(multinom) in the mgcv package
#
# Choose k large enough to allow enough flexibility, the penalty does the rest
# BS is taken the same as the mode used to obtain the model with the smallest GCV
mod <- gam(
  formula = list(y ~ s(age, bs = "gp", k = 15), ~ s(age, bs = "gp", k = 15)),
  family = multinom(K = K - 1),
  data = bmi.data.long,
  method = "REML")

# Show summary of model fit
summary(mod)
```

```{r Predict multinomial P-splines}

# Settings
n.pred   <- length(age.pred)

# Make prediction
# Returns a list of length two (fit and se.fit) containing an n.pred x K matrix
pred <- predict(mod,
                newdata = data.frame(age = age.pred),
                type = "response",
                se.fit = TRUE)

# Rename fit -> E.pi and se.fit -> se.pi
names(pred) <- c("E.pi", "se.pi")

# Add 95% CI
pred <- within(pred, {
  l.pi <- E.pi - qnorm(p = 0.975)*se.pi
  u.pi <- E.pi + qnorm(p = 0.975)*se.pi
})

# This is what we have created
str(pred)


```

```{r Optional: add simulations to pred object. Needed for 95% CI of transitions}

# Number of simulations
n.sim <- 1000

# Extract spline coefficients and covariance matrix of spline coefficients
beta <- coef(mod)
V.beta <- vcov(mod)
n.beta <- length(beta)/(K-1) # Length of coefficient vector for each category

# Generate n.sim realisations for the coefficients
beta.sim <- rmvn(n = n.sim, mu = coef(mod), V = vcov(mod))

# Create model matrix for age.pred based on mod object
X <- predict(mod, newdata = data.frame(age = age.pred), type = "lpmatrix")

# Create empty n.pred x K x n.sim array in pred
pred$pi.sim <- array(0, dim = c(n.pred, K, n.sim))

# For each realisation of beta
for (i in 1:n.sim) {
  # Put beta.sim into block structure for easier computation of linear predictor
  beta.sim.block <- as.matrix(bdiag(split(x = beta.sim[i, ], f = rep(1:(K - 1), each = n.beta))))
  # Compute linear predictor eta. Zero is reference category
  eta <- cbind(0, X %*% beta.sim.block)
  # Compute pi and put it in pred object
  pred$pi.sim[, , i] <- exp(eta)/rowSums(exp(eta))
}


```

```{r Calculate net transitions}

# Function for solving transportation problems using the simplex algorithm
transport.simplex <- function(supply, demand, cost, add.const.mat = NULL, add.const.dir = NULL, add.const.rhs = NULL) {
  # Input:
  # supply = K x 1 vector with supplies
  # demand = K x 1 vector with demands
  # cost   = K x K matrix with shipping costs
  # add.const.mat = matrix with additional constrains
  # add.const.dir = vector with additional directions
  # add.const.rhs = vector with additional right hand sides
  
  # Output:
  # trans = K x K matrix with transitions
  
  # Simplex
  sol <- lp(
    objective.in = as.vector(cost),
    const.mat = rbind(t(kronecker(diag(K), rep(1, K))), t(kronecker(rep(1, K), diag(K))), add.const.mat),
    const.dir = c(rep("=", 2*K), add.const.dir),
    const.rhs = c(supply, demand, add.const.rhs))
  trans <- matrix(sol$solution, K, K, byrow = TRUE)
  
  # Return output
  return(trans)
}

# Cost matrix
cost <- toeplitz(diffinv(1:(K-1)))

# Are simulations available?
exists.sim <- with(pred, exists("pi.sim"))

# Allocate K x K x n.pred-1 arrays in trans and trans.prob list objects for output
trans      <- list(E.trans      = array(NA, dim = c(K, K, n.pred - 1)))
trans.prob <- list(E.trans.prob = array(NA, dim = c(K, K, n.pred - 1)))

# If simulations are available, allocate K x K x n.pred-1 x n.sim arrays for them as well
if (exists.sim) {
  trans$trans.sim           <- array(NA, dim = c(K, K, n.pred - 1, n.sim))
  trans.prob$trans.prob.sim <- array(NA, dim = c(K, K, n.pred - 1, n.sim))
}

# Estimate transitions and transition probabilities
for (i in 1:(n.pred - 1)) {
  # Calculate net transitions and transition probabilities
  trans$E.trans[, , i] <- transport.simplex(
    
    supply = pred$E.pi[i, ],
    demand = pred$E.pi[i + 1, ],
    cost = cost)
  trans.prob$E.trans.prob[, , i] <- t(trans$E.trans[, , i]/pred$E.pi[i, ])
  # If simulations are available, do this n.sim times (takes a while if n.sim is large)
  if (exists.sim) {
    cat("age.pred =", i, "\n")
    for (j in 1:n.sim) {
      trans$trans.sim[, , i, j] <- transport.simplex(
        supply = pred$pi.sim[i, , j],
        demand = pred$pi.sim[i + 1, , j],
        cost = cost)
      trans.prob$trans.prob.sim[, , i, j] <- t(trans$trans.sim[, , i, j]/pred$pi.sim[i, , j])
    }
  }
}

# Calculate summary statistics, only if simulations are available
if (exists.sim) {
  trans <- within(trans, {
    E.trans <- apply(X = trans.sim, MARGIN = c(1, 2, 3), FUN = mean) # Replaces the previous E.trans
    l.trans <- apply(X = trans.sim, MARGIN = c(1, 2, 3), FUN = quantile, prob = 0.025)
    u.trans <- apply(X = trans.sim, MARGIN = c(1, 2, 3), FUN = quantile, prob = 0.975)
  })
  trans.prob <- within(trans.prob, {
    E.trans.prob <- apply(X = trans.prob.sim, MARGIN = c(1, 2, 3), FUN = mean) # Replaces the previous E.trans.prob
    l.trans.prob <- apply(X = trans.prob.sim, MARGIN = c(1, 2, 3), FUN = quantile, prob = 0.025)
    u.trans.prob <- apply(X = trans.prob.sim, MARGIN = c(1, 2, 3), FUN = quantile, prob = 0.975)
  })
}

# This is what we have created
str(trans)
str(trans.prob)
```

```{r Save Transition Probabilities for Women}

trans.prob[["E.trans.prob"]]                 <- provideDimnames(trans.prob[["E.trans.prob"]],
                                                                base = list(c("no","ow","ob"),
                                                                            c("no","ow","ob"),
                                                                            paste("age",age.pred[-n.pred],sep="_")))
dimnames(trans.prob[["E.trans.prob"]])[[3]]  <-  paste("age",age.pred[-n.pred],sep="_")
saveRDS(trans.prob, file= paste(wd,paste("P1matrix_Female_",input_year,".RData",sep=""),sep=""))
```


##03.02 Men, 1-year age BMI prevalences taken from year 2022.

```{r Import Data - Use 2022 only}
df        <- read_csv(paste(wd,"/data/NCD_RisC_Lancet_smooth_2030_Bulgaria.csv",sep = ""))
df        <- df[df$year == input_year & df$sex == "Male",c("age","no","ow","ob")]
df[,2:4]  <- round(df[,2:4] * 1000)
names(df) <- c("age","normal","overweight","obese")
bmi.data  <- df

head(bmi.data)
```

```{r Fit multinomial P-splines}

# Reshape bmi.data into person specific records (long format)
# We need this for the multinomial P-spline fit procedure
# 0 = normal, 1 = overweight, 2 = obese
bmi.data.long <- with(bmi.data, data.frame(
  age = rep(rep(age, times = K), times = c(normal, overweight, obese)),
  y = rep(0:(K - 1), times = c(sum(normal), sum(overweight), sum(obese)))))

# Fit multinomial P-splines
# See help(multinom) in the mgcv package
#
# Choose k large enough to allow enough flexibility, the penalty does the rest
# BS is taken the same as the mode used to obtain the model with the smallest GCV
mod <- gam(
  formula = list(y ~ s(age, bs = "gp", k = 15), ~ s(age, bs = "gp", k = 15)),
  family = multinom(K = K - 1),
  data = bmi.data.long,
  method = "REML")

# Show summary of model fit
summary(mod)
```

```{r Predict multinomial P-splines}

# Make prediction
# Returns a list of length two (fit and se.fit) containing an n.pred x K matrix
pred <- predict(mod,
                newdata = data.frame(age = age.pred),
                type = "response",
                se.fit = TRUE)

# Rename fit -> E.pi and se.fit -> se.pi
names(pred) <- c("E.pi", "se.pi")

# Add 95% CI
pred <- within(pred, {
  l.pi <- E.pi - qnorm(p = 0.975)*se.pi
  u.pi <- E.pi + qnorm(p = 0.975)*se.pi
})

# This is what we have created
str(pred)


```

```{r Optional: add simulations to pred object. Needed for 95% CI of transitions}

# Number of simulations
n.sim <- 1000

# Extract spline coefficients and covariance matrix of spline coefficients
beta <- coef(mod)
V.beta <- vcov(mod)
n.beta <- length(beta)/(K-1) # Length of coefficient vector for each category

# Generate n.sim realisations for the coefficients
beta.sim <- rmvn(n = n.sim, mu = coef(mod), V = vcov(mod))

# Create model matrix for age.pred based on mod object
X <- predict(mod, newdata = data.frame(age = age.pred), type = "lpmatrix")

# Create empty n.pred x K x n.sim array in pred
pred$pi.sim <- array(0, dim = c(n.pred, K, n.sim))

# For each realisation of beta
for (i in 1:n.sim) {
  # Put beta.sim into block structure for easier computation of linear predictor
  beta.sim.block <- as.matrix(bdiag(split(x = beta.sim[i, ], f = rep(1:(K - 1), each = n.beta))))
  # Compute linear predictor eta. Zero is reference category
  eta <- cbind(0, X %*% beta.sim.block)
  # Compute pi and put it in pred object
  pred$pi.sim[, , i] <- exp(eta)/rowSums(exp(eta))
}


```

```{r Calculate net transitions}

# Cost matrix
cost <- toeplitz(diffinv(1:(K-1)))

# Are simulations available?
exists.sim <- with(pred, exists("pi.sim"))

# Allocate K x K x n.pred-1 arrays in trans and trans.prob list objects for output
trans      <- list(E.trans      = array(NA, dim = c(K, K, n.pred - 1)))
trans.prob <- list(E.trans.prob = array(NA, dim = c(K, K, n.pred - 1)))

# If simulations are available, allocate K x K x n.pred-1 x n.sim arrays for them as well
if (exists.sim) {
  trans$trans.sim           <- array(NA, dim = c(K, K, n.pred - 1, n.sim))
  trans.prob$trans.prob.sim <- array(NA, dim = c(K, K, n.pred - 1, n.sim))
}

# Estimate transitions and transition probabilities
for (i in 1:(n.pred - 1)) {
  # Calculate net transitions and transition probabilities
  trans$E.trans[, , i] <- transport.simplex(
    
    supply = pred$E.pi[i, ],
    demand = pred$E.pi[i + 1, ],
    cost = cost)
  trans.prob$E.trans.prob[, , i] <- t(trans$E.trans[, , i]/pred$E.pi[i, ])
  # If simulations are available, do this n.sim times (takes a while if n.sim is large)
  if (exists.sim) {
    cat("age.pred =", i, "\n")
    for (j in 1:n.sim) {
      trans$trans.sim[, , i, j] <- transport.simplex(
        supply = pred$pi.sim[i, , j],
        demand = pred$pi.sim[i + 1, , j],
        cost = cost)
      trans.prob$trans.prob.sim[, , i, j] <- t(trans$trans.sim[, , i, j]/pred$pi.sim[i, , j])
    }
  }
}

# Calculate summary statistics, only if simulations are available
if (exists.sim) {
  trans <- within(trans, {
    E.trans <- apply(X = trans.sim, MARGIN = c(1, 2, 3), FUN = mean) # Replaces the previous E.trans
    l.trans <- apply(X = trans.sim, MARGIN = c(1, 2, 3), FUN = quantile, prob = 0.025)
    u.trans <- apply(X = trans.sim, MARGIN = c(1, 2, 3), FUN = quantile, prob = 0.975)
  })
  trans.prob <- within(trans.prob, {
    E.trans.prob <- apply(X = trans.prob.sim, MARGIN = c(1, 2, 3), FUN = mean) # Replaces the previous E.trans.prob
    l.trans.prob <- apply(X = trans.prob.sim, MARGIN = c(1, 2, 3), FUN = quantile, prob = 0.025)
    u.trans.prob <- apply(X = trans.prob.sim, MARGIN = c(1, 2, 3), FUN = quantile, prob = 0.975)
  })
}

# This is what we have created
str(trans)
str(trans.prob)
```

```{r Save Transition Probabilities for Men}


trans.prob[["E.trans.prob"]] <- provideDimnames(trans.prob[["E.trans.prob"]],base = list(c("no","ow","ob"),
                                                                             c("no","ow","ob"),
                                                                             paste("age",age.pred[-n.pred],sep="_")))
dimnames(trans.prob[["E.trans.prob"]])[[3]]  <-  paste("age",age.pred[-n.pred],sep="_")
saveRDS(trans.prob, file= paste(wd,paste("/P1matrix_Male_",input_year,".RData",sep=""),sep=""))
```

```{r Figures}

trans.prob <- readRDS(paste(wd,paste("P1matrix_Female_",input_year,".RData",sep=""),sep=""))
M_Male     <- readRDS(paste(wd,paste("P1matrix_Male_",input_year,".RData",sep=""),sep=""))

trans.prob[[2]] <- M_Male$E.trans.prob
names(trans.prob) <- c("E.trans.prob","E.trans.prob_M")

# Names states (for titles and legend)
names.states <- c("normal weight", "overweight", "obese")

#
# Plot smoothed prevalences
#

# Set graphical parameters
par(mar = c(4.5, 4.5, 0.5, 0.5))

# Make the plot
plot.new()
plot.window(
  xlim = range(age.pred),
  ylim = c(0, 1))
# Smooth prevalences
with(pred, {
  for (i in 1:K) polygon(
    x = c(age.pred, rev(age.pred)),
    y = c(l.pi[, i], rev(u.pi[, i])),
    col = grey(0.95), border = NA)
  matlines(x = age.pred, y = E.pi, lty = 1, col = 1)
  matlines(x = age.pred, y = l.pi, lty = 3, col = grey(0.5))
  matlines(x = age.pred, y = u.pi, lty = 3, col = grey(0.5))
})
# Add data (as proportions)
with(bmi.data, matpoints(
  x = age,
  y = prop.table(as.matrix(bmi.data[, -1]), margin = 1),
  cex = 0.7, col = 1))
# Axes, titles and legend
axis(side = 1)
axis(side = 2)
box()
title(
  xlab = "Age",
  ylab = "Prevalence")
legend("topright",
       legend = names.states,
       pch = as.character(1:K),
       bty = "n")

#
# Plot transitions in K x K matrix
#

# Set graphical parameters
par(
  mar = c(3, 3, 1.5, 0.1),
  mfrow = c(K, K),
  mgp = c(1.8, 0.7, 0))

# Make the plot
for (i in 1:K) {
  for (j in 1:K) {
    plot.new()
    with(trans, {
      # If simulations are not available
      if (!exists.sim) {
        # Only plot the expected values
        plot.window(
          xlim = range(age.pred),
          ylim = range(E.trans[i, j, ]))
        lines(x = age.pred[-1], y = E.trans[i, j, ], lty = 1, col = 1)
      } else {
        # Plot expected values and confidence intervals
        plot.window(
          xlim = range(age.pred),
          ylim = range(c(l.trans[i, j, ], u.trans[i, j, ])))
        polygon(
          x = c(age.pred[-n.pred], rev(age.pred[-1])),
          y = c(l.trans[i, j, ], rev(u.trans[i, j, ])),
          col = grey(0.95), border = NA)
        lines(x = age.pred[-1], y = l.trans[i, j, ], lty = 3, col = grey(0.5))
        lines(x = age.pred[-1], y = u.trans[i, j, ], lty = 3, col = grey(0.5))
        lines(x = age.pred[-1], y = E.trans[i, j, ], lty = 1, col = 1)
      }
      # Axes and titles
      axis(1)
      axis(2)
      box()
      title(
        main = paste(names.states[i], "to", names.states[j]),
        xlab = "Age",
        ylab = "Transition")
    })
  }
}

#
# Plot transition probabilities in K x K matrix
#

# Set graphical parameters
#par(
 # mar = c(3, 3, 1.5, 0.1),
  #mfrow = c(K, K),
  #mgp = c(1.8, 0.7, 0))

m <- matrix(data = c(1:10,10,10),nrow = 4, ncol = 3, byrow = TRUE)
layout(m, widths = c(3,3,3), heights = c(3,3,3,1))
par(
  mar = c(3, 3, 1.5, 0.1),
  mgp = c(1.8, 0.7, 0))
# Make the plot
for (j in 1:K) {
  for (i in 1:K) {
    plot.new()
    with(trans.prob, {
      # If simulations are not available
      if (!exists.sim) {
        # Only plot the expected values
        plot.window(
          xlim = range(age.pred),
          ylim = range(E.trans.prob_M[i, j, ]))
        lines(x = age.pred[-1], y = E.trans.prob[i, j, ], lty = 1, col = "dark red",lwd = 1)
        lines(x = age.pred[-1], y = E.trans.prob_M[i, j, ], lty = 1, col = "dark blue",lwd = 1)
      } else {
        # Plot expected values and confidence intervals
        plot.window(
          xlim = range(age.pred),
          ylim = range(c(l.trans.prob[i, j, ], u.trans.prob[i, j, ])))
        polygon(
          x = c(age.pred[-n.pred], rev(age.pred[-1])),
          y = c(l.trans.prob[i, j, ], rev(u.trans.prob[i, j, ])),
          col = grey(0.95), border = NA)
        lines(x = age.pred[-1], y = l.trans.prob[i, j, ], lty = 3, col = grey(0.5))
        lines(x = age.pred[-1], y = u.trans.prob[i, j, ], lty = 3, col = grey(0.5))
        lines(x = age.pred[-1], y = E.trans.prob[i, j, ], lty = 1, col = 1)
      }
      # Axes and titles
      axis(side = 1)
      axis(side = 2)
      box()
      title(
        main = paste(names.states[j], "to", names.states[i]),
        cex.main = 0.95,
        xlab = "Age",
        ylab = "Transition probability")
    })
  }
}
par(mar=c(0,0,0,0))
plot(1, type = "n", axes=FALSE, xlab="", ylab="")
legend(x="center",legend = c("Female","Male"),fill = c("dark red","dark blue"),title="Legend", horiz = TRUE)




```

